Modelling low Mach number stellar hydrodynamics with MAESTROeX

### 2019-10-03 astronum_2019_MAESTROeX/astronum-maestroex.tex

The technique that we use in \maestroex~is to modify the fluid equations themselves so as to filter out the soundwaves. This is a similar approach to the incompressible \cite{Boussinesq1901} and anelastic \cite{Ogura1962a,Gough1968,Durran1989} approximations, however the approximation that we use (called the low Mach number approximation \cite{Day2000,Almgren2006a,Nonaka2010} and the generalised pseudo-incompressible approximation \cite{Vasil2013}) allows for background stratification and large density and temperature perturbations due to heating and changes in composition. This is achieved by decomposing the pressure into a 1d hydrostatic base state, $p_0 = p_0(r, t)$, and a dynamic pressure perturbation, $\pi = \pi(r, \bm{x}, t)$, such that the full pressure is given by $p(r, \bm{x}, t) = p_0(r, t) + \pi(\bm{x}, t)$. In the low Mach number regime, asymptotic analysis shows that $|\pi|/p_0 = O(\mathrm{Ma}^2)$.

The low Mach number fluid equations are given by 
\begin{align}
    \pd{\left(\rho X_k\right)}{t} &= - \nabla\cdot\left(\rho X_k \Ub \right) + \rho\dot{\omega}_k, \\
    \pd{\Ub}{t} &= - \Ub\cdot\nabla\Ub - \frac{\beta_0}{\rho}\nabla\left(\frac{\pi}{\beta_0}\right) - \frac{\rho - \rho_0}{\rho}\bm{e}_r,\\
    \pd{\left(\rho h\right)}{t} &= -\nabla\cdot\left(\rho h \Ub\right) + \md{p_0}{t} + \rho H_{\mathrm{nuc}},
\end{align}
where $\rho$, $\Ub$ and $h$ are the mass density, fluid velocity and specific enthalpy, $X_k$ and $\dot{\omega}_k$ are the species mass fraction and production rate of species $k$, and $H_{\mathrm{nuc}}$ is the energy release per time per unit mass. From the base state pressure, we can define a 1d base state density, $\rho_0 = \rho_0(r, t)$, which represents the lateral average and is in hydrostatic equilibrium with $p_0$:
\begin{equation}
    \nabla p_0 = -\rho_0 g \bm{e}_r,
\end{equation}
where $g = g(r,t)$ is the magnitude of the gravitational acceleration, and $\bm{e}_r$ is the radial unit vector. The background stratification is captured by introducing a buoyancy-like term, $\beta_0$. It is defined as 
\begin{equation}
    \beta_0(r,t) = \rho_0(0,t) \exp\left(\int_0^r dr'\, \frac{1}{\overline{\Gamma}_1p_0}\pd{p_0}{r'} \right),
\end{equation}
where $\overline{\Gamma}_1$ is the lateral average of the first adiabatic exponent, $\Gamma_1 \equiv d(\ln p)/d(\ln \rho)|_s$, and $s$ is the entropy. The equations are closed by casting the equation of state (EoS) as a velocity divergence constraint. This is done by taking the Lagrangian derivative of the EoS for pressure as a function of the thermodynamic variables, substituting in the equations of motion for mass and energy, and requiring that the pressure is described by a function of $r$ and $t$ based on the condition of hydrostatic equilibrium. Details of this derivation can be found in \cite{Almgren2006a,Almgren2006b}. This constraint is given by 
\begin{equation}
    \nabla\cdot\left(\beta_0\Ub\right) = \beta_0 \left(S - \frac{1}{\overline{\Gamma}_1 p_0}\pd{p_0}{t} \right).
\end{equation}
Here, $S$ is an expansion term which describes local compressibility effects due to changes in composition and heating from reactions:
\begin{equation}
    S = -\sigma \sum_k \xi_k\dot{\omega}_k + \frac{1}{\rho p_\rho}\sum_k p_{X_k}\dot{\omega}_k + \sigma H_{\mathrm{nuc}},
\end{equation}
where we define the following thermodynamic quantities as
\begin{align*}
    p_{X_k} \equiv \left.\pd{p}{X_k}\right|_{\rho,T,X_{j,j\neq k}},\qquad
     \xi_k&\equiv \left.\pd{h}{X_k}\right|_{p, T,X_{j,j\neq k}},\qquad
     p_\rho\equiv \left.\pd{p}{\rho}\right|_{T, X_k},\\
     \sigma \equiv \frac{p_T}{\rho c_p p_\rho}, \qquad
     p_T&\equiv \left.\pd{p}{T}\right|_{\rho, X_k} \quad\mathrm{and}\quad 
     c_p\equiv \left.\pd{h}{T}\right|_{p, X_k}.
\end{align*}

@@

All codes in the \amrexastro~suite are open source, with all development carried out in public repositories hosted on GitHub\footnote{\url{https://github.com/AMReX-Astro}}. We believe that having both the codes and the development process completely open promotes good scientific practices, as it means that results from our simulations are reproducible and the codes used to produce them can be investigated.\MarginPar{I feel like there is a better word here than investigated} Others can use the codes for their own scientific investigations, and it is possible to adapt all or parts of the codes to suite new problems. Using version control allows us to keep a record of the codes' development process, helping us to track down bugs and means that new and existing developers can learn from previous mistakes. The development branches of the codes are tested nightly using a test suite of problems, checking that any new additions to the code have not significantly changed any of the solutions or slowed down the performance.

@@

As described in \cite{Zingale2011}, to incorporate rotation in our equation set, we add the Coriolis and centrifugal terms to the velocity evolution equation:
\begin{equation}
    \pd{\Ub}{t} = -\Ub\cdot\nabla\Ub - \frac{\beta_0}{\rho}\nabla\left(\frac{\pi}{\beta_0}\right) - \frac{\rho - \rho_0}{\rho}\bm{e}_r - 2\boldsymbol{\Omega} \times \Ub - \boldsymbol{\Omega} \times (\boldsymbol{\Omega} \times \bm{r}),
\end{equation}
where $\boldsymbol{\Omega}$ is the angular velocity, $\bm{F}_\text{Coriolis}=- 2\boldsymbol{\Omega} \times \Ub$ is the Coriolis force and $\bm{F}_\text{centrifugal}= - \boldsymbol{\Omega} \times (\boldsymbol{\Omega} \times \bm{r})$ the centrifugal force.

The simplest way to include rotation is to ignore the effects of the centrifugal force, and only include the Coriolis force. This was done in \maestro, and should be a reasonable approximation for slowly rotating systems. 
% \MarginPar{In Zingale 2011, equation (2) has the Coriolis and centrifugal terms.  Include that equation here?}

@@

In \maestroex, we currently write the 1d base state as a function of the radial coordinate, $r$. For a spherical system, this therefore assumes the base state to be spherically symmetric. However, a star that is rotating at a significant rate will no longer be spherical: it will instead become oblique, bulging around the equator. In order to capture the star's deformation without sacrificing the 1d base state, we can rewrite the base state pressure as a function of the effective potential: $p_0(r)\rightarrow p_0(\Phi_{\text{eff}})$. We can then rewrite the other base state quantities and the system of low Mach number equations as functions of the new coordinates $(\Phi_\text{eff}, \bm{x}, t)$.

The massive stars that we are interested in modelling are believed to be relatively slow rotators\MarginPar{insert some citation/quantify this}, so it is likely that one of the simpler approaches outlined above should be sufficient. However, for modelling e.g.~type I X-ray bursts, it may be that a more sophisticated approach is needed. Type I X-ray bursts are produced by thermonuclear deflagrations in the liquid surface layers of fast rotating neutron stars. These stars typically rotate at frequencies of $\sim$300\textendash600 Hz, a significant fraction of the stars' breakup velocity. This produces non-negligible deformation of the star about the equator, and may cause surface flows with a significant dependence on the meriodional angle. To properly capture the effects of rotation in such a system, it may therefore be necessary to use a two-dimensional base state, with the base state pressure a function of both the radius and the meriodional angle $p_0(r)\rightarrow p_0(r, \theta)$.

@@

% \begin{itemize}
%     \item Why GPUs? Big computers use them.
%     \item GPUs in \amrex: it does everything for you. Also important to note performance portability: GPU-ported AMReX codes still run on other architectures, optimizations for GPU often improve performance on MPI+OpenMP as well.
%     \item GPUs in \maestroex: ported most of the source terms, working on the hydro. 1d base state causing headaches, especially for big problems where we suspect it may not fit in GPU memory.
%     \item Performance plots? Do these at the hackathon?
% \end{itemize}

@@

As mentioned, our work porting \maestroex~to run on GPUs is still ongoing, however in Figure~\ref{fig:gpu_speedup} we show the speedup we have achieved so far for a number of individual functions that have been offloaded to GPU. The plot shows the average execution time of each function, recorded for the 3d \texttt{reacting\_bubble} problem on the OLCF Summit machine. Both tests were run on a single node, consisting of two IBM POWER9 processors (together providing 42 physical CPU cores) and six NVIDIA Volta GPUs. The CPU test used 42 MPI processes, each with 4 OpenMP threads, and the GPU test used 6 MPI processes, each process associated with a single GPU. Note that the plot's $x$-axis scale is logarithmic. As can be seen, large speedups in excess of 30-40$\times$ were achieved for several of the functions when run on the GPU. Once we have finished porting the remaining routines to GPU and further optimised our code, we hope to achieve this sort of speedup for the entire timestep.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{plots/gpu_speedup}
    \caption{Comparison of the performance of functions on the CPU vs the GPU. The plot shows the average execution time for several functions, recorded for a run of the 3d \texttt{reacting\_bubble} problem on Summit. Both tests were performed using a single node, with the CPU test using 42 MPI processes, each with 4 OpenMP threads, and the GPU test using 6 MPI processes, each process associated with a single GPU. Note that the $x$-axis scale is logarithmic. On the right, we have listed the speedup for each function when run on GPUs.}\label{fig:gpu_speedup}
\end{figure}

@@

\maestroex~is an open source code for modelling low Mach number astrophysical flows. It uses the low Mach number approximation, which allows it to use a much larger timestep than conventional compressible schemes by filtering out the soundwaves. Unlike other sound-proof methods, we still retain background stratification and large pressure perturbations due to local compositional changes and heating, both of which are particularly important for modelling atmospheres and burning.

\maestroex~is a new and improved version of our previous code \maestro. It is based on \amrex~(rather than \fboxlib), which allows us to exploit its powerful new solvers and make use of ongoing improvements. \maestroex~is part of the \amrexastro~suite of open source adaptive mesh refinement astrophysical hydrodynamics codes (which also includes \castro~and \nyx). New features developed in other codes in the family can therefore be easily implemented in \maestroex. As development is open, results from our simulations can be reproduced, and monthly archiving to Zenodo ensures software sustainability.

>>>>

The technique that we use in \maestroex~is to modify the fluid equations themselves so as to filter out the soundwaves. This is a similar approach to the incompressible \cite{Boussinesq1901} and anelastic \cite{Ogura1962a,Gough1968,Durran1989} approximations, however the approximation that we use (called the low Mach number approximation \cite{Day2000,Almgren2006a,Nonaka2010} and the generalised pseudo-incompressible approximation \cite{Vasil2013}) allows for background stratification and large density and temperature perturbations due to heating and changes in composition. This is achieved by decomposing the pressure into a one-dimensional hydrostatic base state, $p_0 = p_0(r, t)$, and a dynamic pressure perturbation, $\pi = \pi(\bm{x}, t)$, such that the full pressure is given by $p(\bm{x}, t) = p_0(r, t) + \pi(\bm{x}, t)$. In the low Mach number regime, asymptotic analysis shows that $|\pi|/p_0 = O(\mathrm{Ma}^2)$.

The low Mach number fluid equations are given by 
\begin{align}
    \pd{\left(\rho X_k\right)}{t} &= - \nabla\cdot\left(\rho X_k \Ub \right) + \rho\dot{\omega}_k, \\
    \pd{\Ub}{t} &= - \Ub\cdot\nabla\Ub - \frac{\beta_0}{\rho}\nabla\left(\frac{\pi}{\beta_0}\right) - \frac{\rho - \rho_0}{\rho}\bm{e}_r,\\
    \pd{\left(\rho h\right)}{t} &= -\nabla\cdot\left(\rho h \Ub\right) + \md{p_0}{t} + \rho H_{\mathrm{nuc}},
\end{align}
where $\rho$, $\Ub$ and $h$ are the mass density, fluid velocity and specific enthalpy, $X_k$ and $\dot{\omega}_k$ are the species mass fraction and production rate of species $k$, and $H_{\mathrm{nuc}}$ is the energy release per time per unit mass. From the base state pressure, we can define a one-dimensional base state density, $\rho_0 = \rho_0(r, t)$, which represents the lateral average and is in hydrostatic equilibrium with $p_0$:
\begin{equation}
    \nabla p_0 = -\rho_0 g \bm{e}_r,
\end{equation}
where $g = g(r,t)$ is the magnitude of the gravitational acceleration, and $\bm{e}_r$ is the radial unit vector. The background stratification is captured by introducing a buoyancy-like term, $\beta_0$. It is defined as 
\begin{equation}
    \beta_0(r,t) = \rho_0(0,t) \exp\left(\int_0^r dr'\, \frac{1}{\overline{\Gamma}_1p_0}\pd{p_0}{r'} \right),
\end{equation}
where $\overline{\Gamma}_1$ is the lateral average of the first adiabatic exponent, $\Gamma_1 \equiv d(\ln p)/d(\ln \rho)|_s$, and $s$ is the entropy. The equations are closed by casting the equation of state (EoS) as a velocity divergence constraint. This is done by taking the Lagrangian derivative of the EoS for pressure as a function of the thermodynamic variables, substituting in the equations of motion for mass and energy, and requiring that the pressure is described by a function of $r$ and $t$ based on the condition of hydrostatic equilibrium. Details of this derivation can be found in \cite{Almgren2006a,Almgren2006b}. This constraint is given by 
\begin{equation}
    \nabla\cdot\left(\beta_0\Ub\right) = \beta_0 \left(S - \frac{1}{\overline{\Gamma}_1 p_0}\pd{p_0}{t} \right).
\end{equation}
Here, $S$ is an expansion term which describes local compressibility effects due to changes in composition and heating from reactions:
\begin{equation}
    S = -\sigma \sum_k \xi_k\dot{\omega}_k + \frac{1}{\rho p_\rho}\sum_k p_{X_k}\dot{\omega}_k + \sigma H_{\mathrm{nuc}},
\end{equation}
where we define the following thermodynamic quantities as
\begin{align*}
    p_{X_k} \equiv \left.\pd{p}{X_k}\right|_{\rho,T,X_{j,j\neq k}},\qquad
     \xi_k&\equiv \left.\pd{h}{X_k}\right|_{p, T,X_{j,j\neq k}},\qquad
     p_\rho\equiv \left.\pd{p}{\rho}\right|_{T, X_k},\\
     \sigma \equiv \frac{p_T}{\rho c_p p_\rho}, \qquad
     p_T&\equiv \left.\pd{p}{T}\right|_{\rho, X_k} \quad\mathrm{and}\quad 
     c_p\equiv \left.\pd{h}{T}\right|_{p, X_k}.
\end{align*}

@@

All codes in the \amrexastro~suite are open source, with all development carried out in public repositories hosted on GitHub\footnote{\url{https://github.com/AMReX-Astro}}. We believe that having both the codes and the development process completely open promotes good scientific practices, as it means that results from our simulations are reproducible and the codes used to produce them can be examined and improved by the community.  Others can use the codes for their own scientific investigations, and it is possible to adapt all or parts of the codes to suite new problems. Using version control allows us to keep a record of the codes' development process, helping us to track down bugs and means that new and existing developers can learn from previous mistakes. The development branches of the codes are tested nightly using a test suite of problems, checking that any new additions to the code have not significantly changed any of the solutions or slowed down the performance.

@@

As described in \cite{Zingale2011}, to incorporate rotation in our equation set, we add the Coriolis and/or centrifugal terms to the velocity evolution equation:
\begin{equation}
    \pd{\Ub}{t} = -\Ub\cdot\nabla\Ub - \frac{\beta_0}{\rho}\nabla\left(\frac{\pi}{\beta_0}\right) - \frac{\rho - \rho_0}{\rho}\bm{e}_r - 2\boldsymbol{\Omega} \times \Ub - \boldsymbol{\Omega} \times (\boldsymbol{\Omega} \times \bm{r}),
\end{equation}
where $\boldsymbol{\Omega}$ is the angular velocity, $\bm{F}_\text{Coriolis}=- 2\boldsymbol{\Omega} \times \Ub$ is the Coriolis force and $\bm{F}_\text{centrifugal}= - \boldsymbol{\Omega} \times (\boldsymbol{\Omega} \times \bm{r})$ the centrifugal force.  Note that for slowly rotating systems it is a reasonable approximation to ignore the centrifugal force.

%The simplest way to include rotation is to ignore the effects of the centrifugal force, and only include the Coriolis force. This was done in \maestro, and should be a reasonable approximation for slowly rotating systems. 
% \MarginPar{In Zingale 2011, equation (2) has the Coriolis and centrifugal terms.  Include that equation here?}

@@

In \maestroex, we currently write the one-dimensional base state as a function of the radial coordinate, $r$. For a spherical system, this therefore assumes the base state to be spherically symmetric. However, a star that is rotating at a significant rate will no longer be spherical: it will instead become oblique, bulging around the equator. In order to capture the star's deformation without sacrificing the base state, we can rewrite the base state pressure as a function of the effective potential: $p_0(r)\rightarrow p_0(\Phi_{\text{eff}})$. We can then rewrite the other base state quantities and the system of low Mach number equations as functions of the new coordinates $(\Phi_\text{eff}, \bm{x}, t)$.

The massive stars that we are interested in modelling are believed to be relatively slow rotators\MarginPar{insert some citation/quantify this}, so it is likely that one of the simpler approaches outlined above should be sufficient. However, for modelling e.g.~type I X-ray bursts, it may be that a more sophisticated approach is needed. Type I X-ray bursts are produced by thermonuclear deflagrations in the liquid surface layers of fast rotating neutron stars. These stars typically rotate at frequencies of $\sim$300\textendash600 Hz, a significant fraction of the stars' breakup velocity. This produces non-negligible deformation of the star about the equator, and may cause surface flows with a significant dependence on the meriodional angle. To properly capture the effects of rotation in such a system, it may therefore be necessary to use a two-dimensional base state, with the base state pressure a function of both the radius and the meriodional angle $p_0(r)\rightarrow p_0(r, \theta)$.
\MarginPar{should meriodional be meridional (throughout text)?}

@@

% \begin{itemize}
%     \item Why GPUs? Big computers use them.
%     \item GPUs in \amrex: it does everything for you. Also important to note performance portability: GPU-ported AMReX codes still run on other architectures, optimizations for GPU often improve performance on MPI+OpenMP as well.
%     \item GPUs in \maestroex: ported most of the source terms, working on the hydro. one-dimensional base state causing headaches, especially for big problems where we suspect it may not fit in GPU memory.
%     \item Performance plots? Do these at the hackathon?
% \end{itemize}

@@

As mentioned, our work porting \maestroex~to run on GPUs is still ongoing, however in Figure~\ref{fig:gpu_speedup} we show the speedup we have achieved so far for a number of individual functions that have been offloaded to GPU. The plot shows the average execution time of each function, recorded for the three-dimensional \texttt{reacting\_bubble} problem on the OLCF Summit machine. Both tests were run on a single node, consisting of two IBM POWER9 processors (together providing 42 physical CPU cores) and six NVIDIA Volta GPUs. The CPU test used 42 MPI processes, each with 4 OpenMP threads, and the GPU test used 6 MPI processes, each process associated with a single GPU. Note that the plot's $x$-axis scale is logarithmic. As can be seen, large speedups in excess of 30-40$\times$ were achieved for several of the functions when run on the GPU. Once we have finished porting the remaining routines to GPU and further optimised our code, we hope to achieve this sort of speedup for the entire timestep.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{plots/gpu_speedup}
    \caption{Comparison of the performance of functions on the CPU vs the GPU. The plot shows the average execution time for several functions, recorded for a run of the three-dimensional \texttt{reacting\_bubble} problem on Summit. Both tests were performed using a single node, with the CPU test using 42 MPI processes, each with 4 OpenMP threads, and the GPU test using 6 MPI processes, each process associated with a single GPU. Note that the $x$-axis scale is logarithmic. On the right, we have listed the speedup for each function when run on GPUs.}\label{fig:gpu_speedup}
\end{figure}

@@

\maestroex~is an open source code for modelling low Mach number astrophysical flows. It uses the low Mach number approximation, which allows it to use a much larger timestep than conventional compressible schemes by filtering out the soundwaves. Unlike other sound-proof methods, we still retain background stratification and large density and temperature perturbations due to local compositional changes and heating, both of which are particularly important for modelling atmospheres and burning.

\maestroex~is a new and improved version of our previous code \maestro. It is based on \amrex~(rather than \fboxlib), which allows us to exploit its powerful new solvers and make use of ongoing improvements. \maestroex~is part of the \amrexastro~suite of open source adaptive mesh refinement astrophysical hydrodynamics codes (which also includes \castro~and \nyx). New features developed in other codes in the family can therefore be easily implemented in \maestroex. As development is open, results from our simulations can be reproduced, and monthly archiving to Zenodo ensures software sustainability.\MarginPar{Zenodo?}
